---
title: "06_PCA_and_PCR.Rmd"
author: "Callum Weinberg"
date: "May 19, 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/home/clw/Documents/UCSB/Quarter7/PSTAT 262FE/Final Project/PSTAT_262FE_Final")
```

## Libraries

```{r libraries}
library(pls)
library(kernlab)
library(ggplot2)
library(devtools)
#install_github("vqv/ggbiplot")
library(ggbiplot)
library(lubridate)
library(dplyr)
library(cowplot)
library(car)
library(knitr)
```

## Load and Prepare Data

```{r load_data}
load(file="Intermediate/model_data.Rdata")
load(file = "Intermediate/Ozone_CO2_Combined.Rdata")

# Create Matrices
X_Matrix = model_data[,2:ncol(model_data)]
Y_Matrix = as.matrix(model_data[,1])

# Get Column of Hours
X_hours = hour(Ozone_CO2_Combined$Date.Time)

# Center and Scale Data
X_scale_center = scale(X_Matrix,scale = TRUE, center = TRUE)
```


## PCA using Base R (Stats Package)

```{r pca_prcomp}
## Use prcomp function To Get PC and Projection
# Measure Computational Time - 1000 Iterations
time_prcomp_pca = data.frame(time = rep(NA,1000),method = rep("prcomp",1000))
for (i in 1:1000) {
  start = Sys.time() 
  
  # PR Comp Function. To be consistent with other methods,
  # Use the already scaled and centered version
  prcomp_pca = prcomp(X_scale_center,scale = FALSE, center = FALSE)
  
  # Measure Time
  end = Sys.time()
  time_prcomp_pca[i,1] = end-start
  remove(start,end,prcomp_pca)
}

# Rename/Extract PC and Project Matrices
prcomp_pca = prcomp(X_scale_center,scale = FALSE, center = FALSE)
prcomp_pc = prcomp_pca$x
prcomp_proj = prcomp_pca$rotation
```

## PCA Using Eigenvalue Decomposition

```{r pca_eigen}
## Use eigen Function in R to get PC and Projection
# Measure Computational Time - 1000 iterations
time_eigen_pca = data.frame(time = rep(NA,1000),method = rep("eigen",1000))
for (i in 1:1000) {
  start = Sys.time() 
  
  # Calculate Covariance Matrix of Centered and scaled data
  C = t(X_scale_center)%*%X_scale_center
  
  # Use eigen function to get eigen vectors of C
  # Innerproduct of X and eigenvectors to get principal components
  eigen_pc = X_scale_center%*%eigen(C)$vectors
  
  # Measure Time
  end = Sys.time()
  time_eigen_pca[i,1] = end-start
  remove(start,end,C,eigen_pc)
}

# The projection matrix is just the eigenvectors
# Excluding this from computation time since it 
# is not a necessary step (just renaming values
# For consistency)
C = t(X_scale_center)%*%X_scale_center
eigen_pc = X_scale_center%*%eigen(C)$vectors
eigen_proj = eigen(C)$vectors

# Extract Eigenvalues as well
eigen_C_values = eigen(C)$values
```

## PCA Using Singular Value Decomp

```{r pca_SVD}
## Use svd Function in R to get PC and Projection
# Measure Computational Time - 1000 iterations
time_svd_pca = data.frame(time = rep(NA,1000),method = rep("svd",1000))
for (i in 1:1000) {
  start = Sys.time() 
  
  # Perform SVD Using R Function
  X_svd = svd(X_scale_center)
  
  # Get Principal Compents by m
  svd_pc = X_svd$u%*%diag(X_svd$d)
  
  # Measure Time
  end = Sys.time()
  time_svd_pca[i,1] = end-start
  remove(start,end,X_svd,svd_pc)
}

# Extract Values
X_svd = svd(X_scale_center)
svd_pc = X_svd$u%*%diag(X_svd$d)
svd_singular_values = X_svd$d
svd_eigenvalues = svd_singular_values^2
svd_proj = X_svd$v
```

## PCA Using NIPALS Algorithm

```{r pca_nipals}
# Set Seed for Reproducibility (to randomly set t)
set.seed(42)

## Use Nipals Algorithm in R to get PC and Projection
# Measure Computational Time - 1000 iterations
time_nipals_pca = data.frame(time = rep(NA,1000),method = rep("nipals",1000))
for (j in 1:1000) {
  start = Sys.time() 
  
  # Initialize T and P matrices
  n = nrow(X_scale_center)
  p = ncol(X_scale_center)
  T = matrix(NA, nrow = n, ncol = p)
  P = matrix(NA, nrow = p, ncol = p)
    
  # Initialize Raw data Matrix as scaled, centered matrix
  X = X_scale_center
  
  # Choose Threshold for convergence
  epsilon = .000000001
  # Initialize to something greater than epsilon
  e = 1 
  # Counter to see how many iterations it takes for convergence
  counter = 0
  
  # Step 1: Initialize t
  t_vec = matrix(rnorm(n,0,1),byrow = TRUE)
  
  for (i in 1:ncol(X)){
    # Loop for Component
    while(e > epsilon) {
      
      # Check how many loops occur
      counter = counter + 1
    
      # Step 2: Calculate p vector
      p_prime = t((t(t_vec)%*%X)*(as.numeric(1/(t(t_vec)%*%t_vec))))
      
      # Step 3: Rescale Loading Vector
      p_prime = p_prime*(as.numeric(1/sqrt(t(p_prime)%*%p_prime)))
      
      # Step 4: Regress X onto Normalized Loading Vector
      t_vec_new = X%*%p_prime*(as.numeric(1/t(p_prime)%*%p_prime))
      
      # Step 5: Check for Difference between t_vec and t_vec_new
      e = sum((t_vec - t_vec_new)^2)
      t_vec = t_vec_new
    }
    
    # Step 6: Save Score and Loading vectors and deflate X
    T[,i] = t_vec
    P[,i] = p_prime
    # Deflate X
    X = X - t_vec%*%t(p_prime)
    
    # Reset e
    e = 1
  }
  
  # Measure Time
  end = Sys.time()
  time_nipals_pca[j,1] = end-start
  
  # Extract Values
  nipals_pc = T
  nipals_proj = P
  
  remove(start,end,n,p,T,P,X,epsilon,e,
         counter,p_prime,t_vec,t_vec_new)
}
```


## PCA via Kernlab Package (Linear Kernel)

```{r pca_kernlab}
## Use the kernlab package and linear kernel to get PC
# Measure Computational Time - 1000 iterations
time_kernlab_pca = data.frame(time = rep(NA,1000),method = rep("kernlab",1000)) 
for (i in 1:1000){
  start = Sys.time() 
  
  # Linear Kernel, Same as Regular PCA
  kpca_kernlab_linear = kpca(~.,as.data.frame(X_scale_center),kernel="vanilladot",
              kpar=list(),features=7)
  
  # Measure Time
  end = Sys.time()
  time_kernlab_pca[i,1] = end-start
  remove(start,end,kpca_kernlab_linear)
  
}

# Extract/Rename values
kpca_kernlab_linear = kpca(~.,as.data.frame(X_scale_center),kernel="vanilladot",
            kpar=list(),features=7)
kernlab_pc = pcv(kpca_kernlab_linear)
kernlab_proj = rotated(kpca_kernlab_linear)
```


## Plot Histograms of Time

```{r pca_computational_time}
# Create Dataframe for Plotting
comp_df = rbind(time_prcomp_pca,time_eigen_pca,
                time_svd_pca,time_kernlab_pca,time_nipals_pca)

# Histogram for Single Variable
computational_time_density_plot = 
  ggplot(comp_df, aes(x = log(time), color = method, fill = method)) +
  geom_density(alpha = 0.1) +
  labs(x = "ln(Time) per Iteration", y = "Frequency") +
  theme(text = element_text(size = 20),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.key.width=unit(1,"cm"),
    axis.text.y = element_text(angle=90, hjust=1, size = 14),
    axis.text.x = element_text(size = 14),
    axis.title=element_text(size=14,face="bold"))

## Plot for report
## Commented out for consistency
#png(filename = "Images/PCA_Computational_Time.png", width = 960, height = 480)
computational_time_density_plot
#dev.off()
```

## Confirm Methods produce same Principal Components


```{r pc_confirmation}
# Confirm Same Principal Components
sum(abs(prcomp_pc)-abs(eigen_pc))
sum(abs(prcomp_pc)-abs(svd_pc))
sum(abs(prcomp_pc)-abs(nipals_pc))
# Not the same, needs inspection
sum(abs(prcomp_pc)-abs(xmatrix(kpca_kernlab_linear)))

```

## Scree Plot

```{r pca_scree_plot}
# Percent of Variance Explained
percent_variance_explained = 
  data.frame(PC = seq(1,7,by=1), 
             Percent_Variance = (svd_eigenvalues / sum(svd_eigenvalues)))

# Create Scree Plot
pca_scree = ggplot(data = percent_variance_explained, aes(x = PC, y = Percent_Variance)) +
  geom_point() +
  geom_line() +
  geom_text(aes(label = round(Percent_Variance,4)), size = 3, color = "red", nudge_y = -.05) +
  scale_y_continuous(breaks = seq(0,.7,by = .1)) +
  scale_x_continuous(breaks = seq(1,7,by = 1)) +
  labs(x = "Principal Component", y = "Percent of Variance Explained") +
  theme(text = element_text(size = 20),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.key.width=unit(1,"cm"),
    axis.text.y = element_text(angle=90, hjust=1, size = 10),
    axis.text.x = element_text(size = 14),
    axis.title=element_text(size=14,face="bold"))

## Plot for report
png(filename = "Images/PCA_Scree_Plot.png", width = 720, height = 480)
pca_scree
dev.off()
```


## Biplot of Principal Components

```{r pca_biplot}
## Using ggbiplot package designed for PC object


# Get Hours and Grouped Hours
X_hours_df = data.frame(hours = X_hours) %>%
  mutate(hour_group = ifelse(hours %in% c(21,22,23,0,1,2,3,4,5),"Night",
                             ifelse(hours %in% c(15,16,17,18,19,20),"Evening",
                                    ifelse(hours %in% c(6,7,8,9),"Morning","Mid-day"))))


# Plot 1: Just Loading Vectors
biplot_1_2_plain = ggbiplot(prcomp_pca)

# Plot 2: Loading Vectors and Grouping by time of day
biplot_1_2_groups = ggbiplot(prcomp_pca, labels=X_hours_df$hours, groups = X_hours_df$hour_group)

# Plot 3: Loading vectors with Ellipses for groups
biplot_1_2_groups_ellipses = 
  ggbiplot(prcomp_pca, ellipse=TRUE,  labels=X_hours_df$hours, groups = X_hours_df$hour_group)

# Biplot of 3rd and 4th Principal Components
biplot_3_4_groups_ellipses = ggbiplot(prcomp_pca, choices=c(3,4), ellipse=TRUE,  labels=X_hours_df$hours, groups = X_hours_df$hour_group)


## Plots for Report
png(filename = "Images/biplot_pc_1_2.png", width = 960, height = 480)
plot_grid(biplot_1_2_plain,biplot_1_2_groups, labels = NULL, label_size = 12, ncol = 2, nrow = 1)
dev.off()

png(filename = "Images/biplot_pc_1_2_detailed.png", width = 480, height = 480)
biplot_1_2_groups_ellipses
dev.off

png(filename = "Images/biplot_pc_3_4.png", width = 480, height = 480)
biplot_3_4_groups_ellipses
dev.off

```


## Principal Component Regression

```{r linear_model}
# Illustrate why PCR is Useful Here
model_linear_1 = lm(data = Ozone_CO2_Combined, Sample.Measurement ~ 
                      airport + commercial + elec_prod + industrial + nonroad + onroad + residential)

vif(model_linear_1)

## Alternatively, VIF for PC
# PCR Regression Using All Principal Componets
df_example = data.frame(cbind(Y_Matrix,svd_pc))
model_pc_example = lm(X1 ~ ., data = df)

vif(pcr_all_pc)
```

Fairly clear issues with the linear model and multicollinearity.

```{r pcr_by_pc}
# PCR Regression Using All Principal Componets
pcr_all_pc = lm(Y_Matrix ~ svd_pc - 1)
summary(pcr_all_pc)

# PCR Regression Using 4 Principal Componets
pcr_4_pc = lm(Y_Matrix ~ svd_pc[,c(1:4)] - 1)
summary(pcr_4_pc)

kable(pcr_4_pc$coefficients,pcr_4_pc$)


# PCR Regression Using 2 Principal Componets
#pcr_2_pc = lm(Y_Matrix ~ svd_pc[,c(1:2)] - 1)
#summary(pcr_2_pc)

# PCR Regression Using 1 Principal Componets
#pcr_1_pc = lm(Y_Matrix ~ svd_pc[,c(1)] - 1)
#summary(pcr_1_pc)

#pcr_all_coef = pcr_all_pc$coefficients %*% t(svd_proj)
#as.matrix(X_Matrix)%*%t(pcr_all_coef)
```


```{r pcr_pls_package}
## PCR Regression Using PLS Package
# This portion just illustrates that PCR function
# can be used to achieve the same results
can 
pca_regression_model = V1 ~ airport + commercial + elec_prod + industrial + nonroad + onroad + residential
pca_regression = pcr(pca_regression_model, data = as.data.frame(cbind(Y_Matrix,X_scale_center)))

# Illustrating that the coefficients from
# PCR package are the same as those produces when
# fitting the linear model, after those coefficients
# have been projected onto the subspace
pcr_all_pc$coefficients %*% t(svd_proj)
pca_regression$coefficients[,,7]

```